An enhancer code compendium of early human neuronal and neural crest development

### 2025-05-22 manuscript.tex
Text changes

@@

Genomic enhancer code studies have been limited to specific cell types in model organisms. 
We created a compendium of enhancer codes for all cell types of the human neural tube and neural crest. 
We generated a single-cell multiome atlas of human neural tube organoids and of the head of a four-week-post conception 
human embryo. Using DNA sequence based deep learning - DeepNeuralTube - we identify 15 cell types that are shared 
between the organoid and the human embryo and comprehensively analyse the enhancer code of each individual cell type. 
We confirm the paradigm that enhancers are encoded as heterotypic clusters of 4-5 TF binding sites within one 
nucleosome depleted region, using soft syntactic rules. We also provide counter examples to this rule, with enhancers 
encoded as homotypic clusters and by dimer motifs. Finally, we validate DeepNeuralTube using enhancer reporter assays 
and by interpreting genetic variation associated with facial features.

@@

In this study, we address these limitations by building an enhancer code compendium of cell types and cell states generated by human neural tube organoids, 
a model system for early human neural and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021}, 
in direct comparison to an enhancer compendium we generate in parallel for the head of a four-week post conception human 
embryo. Through these comparisons we assess the enhancer code paradigm and construct an enhancer annotation and code 
resource that is a consensus between the organoid model and the human embryo.\par

@@

Finally, to also investigate the performance of the models on migratory neural crest, 
a cell type for which there are no validated enhancers in the VISTA enhancer browser, 
we selected a genomic region that is specifically accessible in the migratory neural crest (\textbf{Fig. S2d}). 
Both models predict this region to be specific to the migratory neural crest class (\textbf{Fig. S2e})
and nucleotide importance scores reveal TFBS for a SOX monomer, a SOX dimer and TFAP2 (\textbf{Fig. 2i}). 
We tested the enhancer activity of this region using an enhancer-reporter assay in chicken embryos. 
Indeed the enhancer was specifically active in migratory neural crest 24 hours after injection in an HH4 embryo (\textbf{Fig. 2i}).\par

@@

To investigate the enhancer code underlying dorsal-ventral patterning of neural progenitors further, we clustered 
and aligned \textit{de novo} motifs, identified in dorsal-ventral regions, based on pairwise motif similarities. 
Next, we calculated the average contribution for all identified motif instances per cluster and for each topic 
resulting in a dorsal-ventral enhancer code-table (\textbf{Fig. 3f}).
Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ found similar patterns underlying the dorsal-ventral states and the 
importance of the instances of these patterns is highly correlated across both models (0.75 average Pearson correlation; \textbf{Fig. 3f}). 
For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD and RFX instances 
are important (\textbf{Fig. 3f-g}) and the importance of these instances correlates with the expression of respectively \textit{FOXA1/A2/P2/P4}, \textit{TEAD1} and \textit{RFX3/4} in both the organoid and embryo (\textbf{Fig. 3h} and Fig. S3). 
In subsequent more dorsal neural progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig. 3f-h)}); 
PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig. 3f-h} and Fig. S3) and 
ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3f-h} and Fig. S3). 
For all dorsal-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), 
SOX instances are important correlating with the expression of \textit{SOX2/3} (\textbf{Fig. 3f-h} and Fig. S3). 
Similarly, ZEB instances have a negative importance to the accessibility of regions in dorsal-ventral domains except the most ventral one (\textbf{Fig. 3f} and Fig. S3). 
Finally, we calculated the Jaccard index of genomic regions based on the presence of instances of each identified pattern to investigate which patterns often co-occur in the same region (\textbf{Fig. 3i}). 
FOX, TEAD and RFX instances often co-occur in the same genomic region while instances of other patterns tend to be present in separate regions (\textbf{Fig. 3i}).\par

@@

It has been reported that FOXA1/2 is a pioneer factor\cite{iwafuchi-doiPioneerTranscriptionFactor2016, iwafuchiGeneNetworkTransitions2020}. 
Indeed, when centering on FOX instances we observe a strong Tn5 cut-site footprint (\textbf{Fig. S4a}). 
We then asked if there is any difference between regions that are generally bound by FOXA2 in any cell state that expresses this TF and regions, bound by FOXA2, that are specific to the floorplate. 
For this purpose, we intersected FOXA2 ChIP-seq peaks in HepG2 and A549  with floorplate peaks containing FOX instances. 
In the FOXA2 target enhancers of the floorplate that are also bound by FOXA2 in other systems (HepG2 and A549 cell lines), 
we identified a higher number of FOXA2 sites using FIMO\cite{
grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025
} compared to FOXA2 target enhancers that are uniquely accessible in the floorplate cells (\textbf{Fig. S4b}). 
Furthermore, individual FOX instances in general FOXA2 target regions have higher affinity compared to FOXA2 instances 
in specific regions (\textbf{Fig. S4c}). This suggests that FOXA2 by itself can bind to a subset of genomic regions 
independently of the cell type (i.e., indepdent of cooperating TFS), but mainly to regions with multiple sites, and 
high-affinity sites.

@@

In the embryo we identified seven topics corresponding to different AP domains, from anterior to posterior this is Topic 61, Topic 59, Topic 31, Topic 62, Topic 70, Topic 52 and Topic 71 (\textbf{Fig. 4a}). 
We first asked if the AP-program is distinct from the DV program. 
In other words, is AP patterning encoded by the same or different genomic regions as DV patterning? 
For this purpose, we calculated the pairwise correlation between region-topic probabilities of all AP and DV topics (\textbf{Fig. 4b}). 
This revealed that both AP and DV patterning are mostly encoded in distinct genomic regions, 
with the exception of AP Topic 59 and 31 that correlate with the region-topic probability of DV Topic 38 
(which corresponds to cells expressing \textit{NKX2-1/2}; \textbf{Fig. 4b}).\par

$DeepNeuralTube^e$ was able to distinguish the different AP domains but grouped topics of similar domains
% TODO: S.A.: How does the organoid model score all these AP topics?
together (i.e., Topic 59, 31 and 62; Topic 52 and 71), resulting in four distinct AP domains (\textbf{Fig.4 c-d}) and 
for this it made use of four distinct homeodomain TF binding motifs, one for each domain (\textbf{Fig. 4e}). 
From anterior to posterior this is \textit{TAATTAG}, \textit{TAATCCCT}, \textit{TCATGWTGANTGA} and \textit{CATYMATCA}.
Notably, the patterns corresponding to the two posterior most topics correspond to homeobox dimer-motifs. 
Previously it had been hypothesized that genome-wide specific binding of homeodomain TFs is accomplished through
cooperative binding of two TFs\cite{mannChapter3Hox2009}. 
Indeed, the presence of instances of the \textit{TCATGWTGANTGA} motif correlates with the expression of both \textit{PAX5/8} and \textit{EN1/2}; 
and the presence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig. 4d-e} and Fig. S5). 
In contrast, patterns corresponding to the two anterior most topics correspond to monomer motifs, 
correlating with the expression of \textit{LHX2/9}, \textit{RAX} and \textit{SIX3/6} for the most anterior topic and 
with the expression of \textit{OTX1/2}, \textit{EMX2}, \textit{DMBX1} and \textit{LMX1A} for the second most anterior
topic (\textbf{Fig. 4d-e} and Fig. S5)\par
%TODO: I'm surprised to see EMX2 (forebrain) together with OTX1/2 and DMBX1 (midbrain)

@@

Organoid Topic 62 and embryo Topic 103 represent the pre-migratory neural crest state and 
instances of TEAD, GRHL, AP2 and ZEB patterns are important for this state (\textbf{Fig. 5b-c}). 
The importance of TEAD instances correlates with \textit{TEAD3/4} expression, GRHL with \textit{GRHL1/1} 
and AP2 with both \textit{TFAP2A} and \textit{TFAP2C} (\textbf{Fig. 5d} and Fig. S6). 
Note that ZEB instances have a positive contribution to the pre-migratory while they have a
negative contribution to all other neural crest states (\textbf{Fig. 5b-c}); 
and moreover, a negative contribution to all other cell states of both the neural progenitors (\textbf{Fig. 3e}) and neurons (\textbf{Fig. 6b}). 
The positive contribution of ZEB in the pre-migratory neural crest state coincides with the absence of the expression of both \textit{ZEB1} and \textit{ZEB2} (\textbf{Fig. 5d} and Fig. S6). 
This is in line with the previously reported role of ZEB as a chromatin closing repressor \cite{taskiranCelltypedirectedDesignSynthetic2024}.\par

@@

Recently, a genome wide association study (GWAS) identified 203 single-nucleotide polymorphisms (SNPs)
that are associated with facial variation\cite{whiteInsightsGeneticArchitecture2021}. 
Given that regions surrounding these SNPs were enriched for enhancer activity in cranial neural crest cells\cite{whiteInsightsGeneticArchitecture2021} 
we asked whether $DeepNeuralTube$ could explain the effect of some of these variants on enhancer activity. For all suggestive SNPs we scored both the reference and alternative allele(s) using both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ 
and calculated the difference in prediction score between both alleles (delta prediction score). 
At an absolute delta prediction score threshold of 0.2 we identified 235 SNPs that are explainable by either of the two models. 
Note that only very few (1 out of 203) of these "explainable" SNPs overlap with the 203 lead SNPs identified by White \textit{et al.} (\textbf{Fig. 5f}). 
Most of the "explainable" SNPs have a high delta prediction score for either organoid neural crest Topic 58 or Topic 59 (\textbf{Fig. S7}). 
As an example, we show the explanation of rs1555067 that is located upstream of \textit{PRRX1}. 
The G \textgreater A substitution of this SNP causes a delta prediction score of 0.64 for organoid Topic 58
(\textbf{Fig. 5g}), increasing the predicted chromatin accessibility of this genomic region in the facial mesenchyme. 
This predicted increase in chromatin accessibility is caused by generating an EBOX site 6bp downstream of a partial
homeobox site potentially creating a new binding site for TWIST1-homeodomain TF dimer (\textbf{Fig. 5h}).\par

@@

For both the organoid and embryo we found two topics (organoid Topic 6 and 4 and embryo Topic 10 and 18) for which
instances of patterns from neural progenitors are important (\textbf{Fig. 6b-c}).
These include instances of TEAD, FOX and RFX (\textbf{Fig. 6b-c}) 
correlating with the expression of respectively \textit{TEAD1},
\textit{FOXA2/J3/O1/P1/P2/P4} and \textit{RFX3/4} (\textbf{Fig. 6d} and Fig. S8). 
Instances of a SOX monomer motif is important for one of these topics (\textbf{Fig. 6b-c}), 
correlating with the expression of \textit{SOX2} (\textbf{Fig. 6d} and Fig. S8). 
Next, in both organoids and embryo cells we identified a topic for which both instances of neural progenitor 
patterns as well as an EBOX pattern is important (organoid Topic 23 and embryo Topic 13 \textbf{Fig. 6b-c}). 
The importance of this EBOX pattern correlates with the expression of \textit{ASCL1} and \textit{NEUROD4}. 
For subsequent topics SOX instances are not important anymore (\textbf{Fig. 6b-c}); 
instead we find topics where a combination of EBOX and EBF instances are important 
(organoid and embryo Topic 24; \textbf{Fig. 6b-c}) the latter correlating with the expression of \textit{EBF1-3}
(\textbf{Fig. 6d} and Fig. S8) 
and topics where a combination of EBF and ONECUT instances are important 
(organoid Topic 13 and embryo Topic 18; \textbf{Fig. 6b-c}), 
the latter correlating with the expression of \textit{ONECUT1-3}. 
Additionally, in the embryo we found topics where instances of NKX are important. 
The importance of these instances is anti-correlated with the expression \textit{NKX2-2} 
and correlated with the expression of \textit{ISL1} (\textbf{Fig. 6d} and Fig. S8). 
Finally, in the organoids we could identify an additional topic with instances of GATA
(\textbf{Fig. 6b-c}), correlating with the expression of \textit{GATA2} (\textbf{Fig. 6d} and Fig. S8).\par
%TODO: C.M.: NKX2-2 is an important patterning factor in the neural tube, specifically for ventralization so it should only be expressed around the floor plate. That you don't see it in the organoid might indicate that part of this patterning is not present. https://www.nature.com/articles/19315
%TODO: S.D.W.: I do see it in the DV patterning part, just not in the subset of differentiating neurons.
%TODO: C.M.: That's interesting, so in the organoid it doesn't seem to persist into neuronal differentiation while the embryo retains that NKX2.2 DV patterning? Are all the same types of neurons there or might it be related to a missing cell type?

@@

@@

that could be important for cell type specific binding.\par

In conclusion, we provide an enhancer code compendium of early human neuronal and neural crest development. 
We confirm the paradigm that most human enhancers consist of heterotypic clusters of TFBS with soft syntactic rules 
but also provide counter exampled to this paradigm. 
Furthermore, we validate the biological relevance of neural tube organoids, 
having representative transcriptomes and genomic regulatory code activation compared to embryo development. 
Finally, we provide $DeepNeuralTubes$, two sequence-to-chromatin accessibility models, 
freely available to the research community, 
that can be used to predict, 
interpret and design enhancers underlying early human neuronal and neural crest development.
Alongside these models, we implemented new tools in our genomic deep learning framework CREsted\cite{kempynckCREstedModelingGenomic2025}, specifically for the analysis of compendia of enhancer codes underlying complex biological systems.

@@

Raw sequencing data, scATAC-seq fragment files, scRNA-seq count matrix, cell metadata, consensus peaks, and cell- and 
region topic probabilities for the human neural tube organoids are available on Gene Expression Omnibus (GEO) using 
accession number \href{https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE295337}{GSE295337}
(reviewer token: ). %TODO: add before submission.
Raw sequencing data of the human embryo will be made available through the European Genome Phenome Archive. %TODO: Add link to EGA
scATAC-seq fragment files, scRNA-seq count matrix, cell metadata, consensus peaks and cell-and region topic
probabilities of the human embryos are available through Zenodo (\href{https://doi.org/10.5281/zenodo.15281826}{10.5281/zenodo.15281826}). 
$DeepNeuralTube_O$ and $DeepNeural$- $Tube_E$ are available via Zenodo (\href{https://doi.org/10.5281/zenodo.15280773}{10.5281/zenodo.15280773}). 
Data can be interactively explored using the \href{https://scope.aertslab.org/#/Neural_Tube/*/welcome}{neural tube SCope session} and on the UCSC genome browser: 
\begin{itemize}
  \item \href{https://genome.ucsc.edu/s/Seppe%20De%20Winter/neural%20tube%20doro%2Dventral}{dorsal-ventral UCSC session}
  \item \href{https://genome.ucsc.edu/s/Seppe%20De%20Winter/neural%20tube%20neural%20crest}{neural crest UCSC session}
  \item \href{https://genome.ucsc.edu/s/Seppe%20De%20Winter/neural%20tube%20neuron}{neuron UCSC session}
  \item \href{https://genome.ucsc.edu/s/Seppe%20De%20Winter/neural%20tube%20anterior%2Dposterior}{anterior-posterior
      UCSC session}
\end{itemize}

@@

We thank Peter L\"onnerberg for preprocessing the human embryo data and making it available through EGA.
We also thank Prof. Dr. Tatjana Sauka-Spengler and Dr. Ruth Williams for their great support with helping setting up the chicken embryo enhancer reporter assays.
Computing was performed at the Vlaamse Supercomputer Center and VIB datacore. 
This work was funded by the following grants to 
S. Aerts , %TODO: Add funding
PhD fellowship from FWO to S. De Winter (1191323N) and 
Post doctoral fellowship from ... to C. Mannens (), %TODO: add funding

>>>>

% set T1 font encoding so underscores are rendered correctly
\usepackage[T1]{fontenc}

@@

Genomic enhancer code studies have been limited to specific cell types in model organisms. 
We created a compendium of enhancer codes for all cell types of the human neural tube and neural crest. 
We generated a single-cell multiome atlas of human neural tube organoids and of the head of a four-week post conception 
human embryo. Using DNA sequence based deep learning - DeepNeuralTube - we identify 15 cell types that are shared 
between the organoid and the human embryo and comprehensively analyse the enhancer code of each individual cell type. 
We confirm the paradigm that enhancers are encoded as heterotypic clusters of 4-5 TF binding sites within one 
nucleosome depleted region, using soft syntactic rules. We also provide counter examples to this rule, with enhancers 
encoded as homotypic clusters and by dimer motifs. Finally, we validate DeepNeuralTube using enhancer reporter assays 
and by interpreting genetic variation associated with facial features.

@@

In this study, we address these limitations by building an enhancer code compendium of cell types and cell states generated by human neural tube organoids, 
a model system for early human neuronal and neural crest development\cite{abdelfattahActuationEnhancesPatterning2021}, 
in direct comparison to an enhancer compendium we generate in parallel for the head of a four-week post conception human 
embryo. Through these comparisons we assess the enhancer code paradigm and construct an enhancer annotation and code 
resource that is a consensus between the organoid model and the human embryo.\par

@@

Finally, to also investigate the performance of the models on migratory neural crest, 
a cell type for which there are no validated enhancers in the VISTA enhancer browser, 
we selected a genomic region that is specifically accessible in the migratory neural crest (\textbf{Fig. S2d}). 
Both models predict this region to be specific to the migratory neural crest class (\textbf{Fig. S2e})
and nucleotide importance scores reveal TFBS for a SOX monomer, a SOX dimer and AP2 (\textbf{Fig. 2i}). 
We tested the enhancer activity of this region using an enhancer-reporter assay in chicken embryos. 
Indeed the enhancer was specifically active in migratory neural crest 24 hours after injection in an HH4 embryo (\textbf{Fig. 2i}).\par

@@

To investigate the enhancer code underlying dorsal-ventral patterning of neural progenitors further, we clustered 
and aligned \textit{de novo} motifs, identified in dorsal-ventral regions, based on pairwise motif similarities. 
Next, we calculated the average contribution for all identified motif instances per cluster and for each topic 
resulting in a dorsal-ventral enhancer code-table (\textbf{Fig. 3f}).
Both $DeepNeuralTube^o$ and $DeepNeuralTube^e$ found similar patterns underlying the dorsal-ventral states and the 
importance of the instances of these patterns is highly correlated across both models (0.75 average Pearson correlation; \textbf{Fig. 3f}). 
For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD and RFX instances 
are important (\textbf{Fig. 3f-g}) and the importance of these instances correlates with the expression of respectively \textit{FOXA1/A2/P2/P4}, \textit{TEAD1} and \textit{RFX3/4} in both the organoid and embryo (\textbf{Fig. 3h} and Fig. S3). 
In subsequent more dorsal neural progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{Fig. 3f-h}); 
PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig. 3f-h} and Fig. S3) and 
ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3f-h} and Fig. S3). 
For all dorsal-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), 
SOX instances are important correlating with the expression of \textit{SOX2/3} (\textbf{Fig. 3f-h} and Fig. S3). 
Similarly, ZEB instances have a negative importance to the accessibility of regions in dorsal-ventral domains except the most ventral one (\textbf{Fig. 3f} and Fig. S3). 
Finally, we calculated the Jaccard index of genomic regions based on the presence of instances of each identified pattern to investigate which patterns often co-occur in the same region (\textbf{Fig. 3i}). 
FOX, TEAD and RFX instances often co-occur in the same genomic region while instances of other patterns tend to be present in separate regions (\textbf{Fig. 3i}).\par

@@

It has been reported that FOXA1/2 is a pioneer factor\cite{iwafuchi-doiPioneerTranscriptionFactor2016, iwafuchiGeneNetworkTransitions2020}. 
Indeed, when centering on FOX instances we observe a strong Tn5 cut-site footprint (\textbf{Fig. S4a}). 
We then asked if there is any difference between regions that are generally bound by FOXA2 in any cell state that expresses this TF and regions, bound by FOXA2, that are specific to the floorplate. 
For this purpose, we intersected FOXA2 ChIP-seq peaks in HepG2 and A549  with floorplate peaks containing FOX instances. 
In the FOXA2 target enhancers of the floorplate that are also bound by FOXA2 in other systems (HepG2 and A549 cell lines), 
we identified a higher number of FOXA2 sites using FIMO\cite{
grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025} compared to FOXA2 target enhancers that are uniquely accessible in the floorplate cells (\textbf{Fig. S4b}). 
Furthermore, individual FOX instances in general FOXA2 target regions have higher affinity compared to FOXA2 instances 
in specific regions (\textbf{Fig. S4c}). This suggests that FOXA2 by itself can bind to a subset of genomic regions 
independently of the cell type (i.e., independent of cooperating TFs), but mainly to regions with multiple sites, and 
high-affinity sites.

@@

In the embryo we identified seven topics corresponding to different AP domains, from anterior to posterior this is Topic 61, Topic 59, Topic 31, Topic 62, Topic 70, Topic 52 and Topic 71 (\textbf{Fig. 4a}). 
We first asked if the AP-program is distinct from the DV-program. 
In other words, is AP patterning encoded by the same or different genomic regions as DV patterning? 
For this purpose, we calculated the pairwise correlation between region-topic probabilities of all AP and DV topics (\textbf{Fig. 4b}). 
This revealed that both AP and DV patterning are mostly encoded in distinct genomic regions, 
with the exception of AP Topic 59 and 31 that correlate with the region-topic probability of DV Topic 38 
(which corresponds to cells expressing \textit{NKX2-1/2}; \textbf{Fig. 4b}).\par

$DeepNeuralTube^e$ was able to distinguish the different AP domains but grouped topics of similar domains
% TODO: S.A.: How does the organoid model score all these AP topics?
together (i.e., Topic 59, 31 and 62; Topic 52 and 71), resulting in four distinct AP domains (\textbf{Fig.4 c-d}) and 
for this it made use of four distinct homeodomain TF binding motifs, one for each domain (\textbf{Fig. 4e}). 
From anterior to posterior this is \textit{TAATTAG}, \textit{TAATCCCT}, \textit{TCATGWTGANTGA} and \textit{CATYMATCA}.
Previously it had been hypothesized that genome-wide specific binding of homeodomain TFs is accomplished through
cooperative TF-binding\cite{mannChapter3Hox2009}. 
Indeed, the presence of instances of the \textit{TCATGWTGANTGA} motif correlates with the expression of both \textit{PAX5/8} and \textit{EN1/2}; 
and the presence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig. 4d-e} and Fig. S5), which resembles cooperative binding of Hox proteins 
in \textit{Drosophila}\cite{mannChapter3Hox2009}.
In contrast, patterns corresponding to the two anterior most topics correspond to monomer motifs, 
correlating with the expression of \textit{LHX2/9}, \textit{RAX} and \textit{SIX3/6} for the most anterior topic and 
with the expression of \textit{OTX1/2}, \textit{EMX2}, \textit{DMBX1} and \textit{LMX1A} for the second most anterior
topic (\textbf{Fig. 4d-e} and Fig. S5). The binding specificity of these TFs might be modulated by another co-factor, 
analogous to Hox-Exd co-binding in fly (termed "latent specificity")\cite{slatteryCofactorBindingEvokes2011}. PBX1-3 are the human orthologs of Exd.\par
%TODO: I'm surprised to see EMX2 (forebrain) together with OTX1/2 and DMBX1 (midbrain)

@@

Organoid Topic 62 and embryo Topic 103 represent the pre-migratory neural crest state and 
instances of TEAD, GRHL, AP2 and ZEB patterns are important for this state (\textbf{Fig. 5b-c}). 
The importance of TEAD instances correlates with \textit{TEAD3/4} expression, GRHL with \textit{GRHL1/2} 
and AP2 with both \textit{TFAP2A} and \textit{TFAP2C} (\textbf{Fig. 5d} and Fig. S6). 
Note that ZEB instances have a positive contribution to the pre-migratory while they have a
negative contribution to all other neural crest states (\textbf{Fig. 5b-c}); 
and moreover, a negative contribution to all other cell states of both the neural progenitors (\textbf{Fig. 3e}) and neurons (\textbf{Fig. 6b}). 
The positive contribution of ZEB in the pre-migratory neural crest state coincides with the absence of the expression of both \textit{ZEB1} and \textit{ZEB2} (\textbf{Fig. 5d} and Fig. S6). 
This is in line with the previously reported role of ZEB as a chromatin closing repressor \cite{taskiranCelltypedirectedDesignSynthetic2024}.\par

@@

Recently, a genome wide association study (GWAS) identified single-nucleotide polymorphisms (SNPs) that are associated 
with facial variation\cite{whiteInsightsGeneticArchitecture2021}. 
Given that regions surrounding these SNPs were enriched for enhancer activity in cranial neural crest cells\cite{whiteInsightsGeneticArchitecture2021} 
we asked whether $DeepNeuralTube$ could explain the effect of some of these variants. 
We scored both the reference and alternative allele(s) using both $DeepNeuralTube$ models and calculated the difference 
in prediction score between alleles (delta prediction score). 
At an absolute delta prediction score threshold of 0.2 we identified 235 SNPs, out a total of 23,701 SNPs associated 
with facial variation, that are explainable by either of the two models (\textbf{Fig. 5f}). 
Most of the "explainable" SNPs have a high delta prediction score for either the migratory neural crest class 
(organoid Topic 59) or the facial mesenchyme class (organoid Topic 58; \textbf{Fig. S7}).
As an example, we show the explanation of rs1555067 that is located upstream of \textit{PRRX1}. 
The G \textgreater A substitution of this SNP causes a delta prediction score of 0.64 for organoid Topic 58
(\textbf{Fig. 5g}), increasing the predicted chromatin accessibility of this genomic region in the facial mesenchyme. 
This predicted increase in chromatin accessibility is caused by generating an EBOX site 6bp downstream of a partial
homeobox site potentially creating a new binding site for TWIST1-homeodomain TF dimer (\textbf{Fig. 5h}).\par

@@

For both the organoid and embryo we found two topics (organoid Topic 6 and 4 and embryo Topic 10 and 18) for which
instances of patterns from neural progenitors are important (\textbf{Fig. 6b-c}).
These include instances of TEAD, FOX and RFX (\textbf{Fig. 6b-c}) 
correlating with the expression of respectively \textit{TEAD1},
\textit{FOXA2/J3/O1/P1/P2/P4} and \textit{RFX3/4} (\textbf{Fig. 6d} and Fig. S8). 
Instances of a SOX monomer motif is important for one of these topics (\textbf{Fig. 6b-c}), 
correlating with the expression of \textit{SOX2} (\textbf{Fig. 6d} and Fig. S8). 
Next, in both organoids and embryo cells we identified a topic for which both instances of neural progenitor 
patterns as well as an EBOX pattern is important (organoid Topic 23 and embryo Topic 13 \textbf{Fig. 6b-c}). 
The importance of this EBOX pattern correlates with the expression of \textit{ASCL1} and \textit{NEUROD4}. 
For subsequent topics SOX instances are not important anymore (\textbf{Fig. 6b-c}); 
instead we find topics where a combination of EBOX and EBF instances are important 
(organoid and embryo Topic 24; \textbf{Fig. 6b-c}) the latter correlating with the expression of \textit{EBF1-3}
(\textbf{Fig. 6d} and Fig. S8) 
and topics where a combination of EBF and ONECUT instances are important 
(organoid Topic 13 and embryo Topic 18; \textbf{Fig. 6b-c}), 
the latter correlating with the expression of \textit{ONECUT1-3}. 
Additionally, in the embryo we found topics where instances of NKX are important. 
The importance of these instances is anti-correlated with the expression \textit{NKX2-2} 
and correlated with the expression of \textit{ISL1} (\textbf{Fig. 6d} and Fig. S8). 
Finally, in the organoids we could identify an additional topic with instances of GATA
(\textbf{Fig. 6b-c}), correlating with the expression of \textit{GATA2} (\textbf{Fig. 6d} and Fig. S8), a marker 
gene for the induction of V2 interneurons\cite{zhouGATA2RequiredGeneration2000, delbarrioRegulatoryNetworkInvolving2007,
sagnerEstablishingNeuronalDiversity2019}.\par
%TODO: C.M.: NKX2-2 is an important patterning factor in the neural tube, specifically for ventralization so it should only be expressed around the floor plate. That you don't see it in the organoid might indicate that part of this patterning is not present. https://www.nature.com/articles/19315
%TODO: S.D.W.: I do see it in the DV patterning part, just not in the subset of differentiating neurons.
%TODO: C.M.: That's interesting, so in the organoid it doesn't seem to persist into neuronal differentiation while the embryo retains that NKX2.2 DV patterning? Are all the same types of neurons there or might it be related to a missing cell type?

@@

This observation is similar to Hox-Exd co-binding, where sequence specificity of HOX proteins is modulated by another 
co-factor\cite{slatteryCofactorBindingEvokes2011}.

@@

that could be important for cell type-specific binding.\par

In conclusion, we provide an enhancer code compendium of early human neuronal and neural crest development. 
We confirm the paradigm that most human enhancers consist of heterotypic clusters of TFBS with soft syntactic rules 
but also provide counter example to this paradigm. 
Furthermore, we validate the biological relevance of neural tube organoids, 
having representative transcriptomes and genomic regulatory code activation compared to embryo development. 
Finally, we provide $DeepNeuralTubes$, two sequence-to-chromatin accessibility models, 
freely available to the research community, 
that can be used to predict, 
interpret and design enhancers underlying early human neuronal and neural crest development.
Alongside these models, we implemented new tools in our genomic deep learning framework CREsted\cite{kempynckCREstedModelingGenomic2025}, specifically for the analysis of compendia of enhancer codes underlying complex biological systems.

@@

Raw sequencing data, scATAC-seq fragment files, scRNA-seq count matrix, cell metadata, consensus peaks, and cell- and 
region topic probabilities for the human neural tube organoids are available on Gene Expression Omnibus (GEO) using 
accession number \href{https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE295337}{GSE295337}
(reviewer token: ). %TODO: add before submission.
Raw sequencing data of the human embryo will be made available through the European Genome Phenome Archive. %TODO: Add link to EGA
scATAC-seq fragment files, scRNA-seq count matrix, cell metadata, consensus peaks and cell-and region topic
probabilities of the human embryos are available through Zenodo (\href{https://doi.org/10.5281/zenodo.15281826}{10.5281/zenodo.15281826}). 
$DeepNeuralTube^O$ and $DeepNeural$- $Tube^E$ are available via Zenodo (\href{https://doi.org/10.5281/zenodo.15280773}{10.5281/zenodo.15280773}). 
Data can be interactively explored using the \href{https://scope.aertslab.org/#/Neural_Tube/*/welcome}{neural tube SCope session} and on the UCSC genome browser: 
\begin{itemize}
  \item \href{https://genome.ucsc.edu/s/Seppe%20De%20Winter/neural%20tube%20doro%2Dventral}{dorsal-ventral UCSC
    session}: https://genome.ucsc.edu/s/Seppe\%20De\%20Winter/neural- \%20tube\%20doro\%2Dventral
  \item \href{https://genome.ucsc.edu/s/Seppe%20De%20Winter/neural%20tube%20neural%20crest}{neural crest UCSC session}: https://genome.ucsc.edu/s/Seppe\%20De\%20Winter/neural- \\\%20tube\%20neural\%20crest
  \item \href{https://genome.ucsc.edu/s/Seppe%20De%20Winter/neural%20tube%20neuron}{neuron UCSC session}: https://genome.ucsc.edu/s/Seppe\%20De\%20Winter/neural\%20- tube\%20neuron
  \item \href{https://genome.ucsc.edu/s/Seppe%20De%20Winter/neural%20tube%20anterior%2Dposterior}{anterior-posterior
      UCSC session}: https://genome.ucsc.edu/s/Seppe\%20De\%20Winter/neur- al\%20tube\%20anterior\%2Dposterior
\end{itemize}

@@

We thank Peter L\"onnerberg for preprocessing the human embryo data and making it available through EGA.
We also thank Prof. Dr. Tatjana Sauka-Spengler and Dr. Ruth Williams for their great support with helping to set up the 
chicken embryo enhancer reporter assays.
Computing was performed at the Vlaamse Supercomputer Center and VIB datacore. 
This work was funded by the following grants to S. Aerts: ERCAdG (101054387); CZI (DI2-0000000068); SBO (S005024N); FWO 
(G0I2722N EOS ID 40007513, G094121N, G044124N); and Foundation Against Cancer (2020-1396 \& 2024-140), 
the following grants to S. Linnarson: Erling-Persson Foundation Human Developmental Cell Atlas grant, Knut and Alice
Wallenberg Foundation grants 2015.0041, 2018.0172 and 2018.0220, Swedish Foundation for Strategic Research SB16-0065, 
and EU Horizon2020 BRAINTIME project 874606 and 
PhD fellowship from FWO to S. De Winter (1191323N).
